---
title: "Analyze ED Outcomes Data"
author: "Zachary del Rosario"
date: 2019-11-24
output:
  github_document
---

On November 20th, the Department of Education released data on [college student
outcomes](https://www.insidehighered.com/news/2019/11/21/federal-government-releases-earnings-data-thousands-college-programs),
for the first time disaggregated by degree program. Here I carry out an
exploratory analysis on the data.

```{r setup}
library(tidyverse)
library(ggrepel)
library(rpnd)
```

```{r load}
file_data <- "./FieldOfStudyData1516_1617_PP.csv"

df_raw <-
  file_data %>%
  read_csv(
    col_types = cols(
      UNITID          = col_double(),
      OPEID6          = col_character(),
      INSTNM          = col_character(),
      CONTROL         = col_character(),
      MAIN            = col_double(),
      CIPCODE         = col_character(),
      CIPDESC         = col_character(),
      CREDLEV         = col_double(),
      CREDDESC        = col_character(),

      COUNT           = col_double(),
      DEBTMEDIAN      = col_double(),
      DEBTPAYMENT10YR = col_double(),
      DEBTMEAN        = col_double(),
      TITLEIVCOUNT    = col_double(),
      EARNINGSCOUNT   = col_double(),
      MD_EARN_WNE     = col_double(),
      IPEDSCOUNT1     = col_character(),
      IPEDSCOUNT2     = col_character()
    )
  )

df_data <-
  df_raw %>%
  select(
    uid               = UNITID,
    institution       = INSTNM,
    control           = CONTROL,
    major             = CIPDESC,
    level             = CREDLEV,
    degree            = CREDDESC,
    count             = COUNT,
    debt_median       = DEBTMEDIAN,
    debt_payment_10yr = DEBTPAYMENT10YR,
    debt_mean         = DEBTMEAN,
    title_iv_count    = TITLEIVCOUNT,
    earnings_count    = EARNINGSCOUNT,
    earnings_median   = MD_EARN_WNE
  ) %>%
  mutate(
    debt_annual = 12 * debt_payment_10yr,
    count = if_else(is.na(count), 0, count),
    earnings_count = if_else(is.na(earnings_count), 0, earnings_count),
    title_iv_count = if_else(is.na(title_iv_count), 0, title_iv_count)
  )
```

## UG Engineering Colleges
<!-- -------------------------------------------------- -->

I have a particular interest in small undergraduate-focused engineering
colleges. How do they compare?

```{r glimpse-ug-engr}
df_data %>%
  filter(
    str_detect(institution, "Olin") |
    str_detect(institution, "Harvey Mudd") |
    str_detect(institution, "Rose-Hulman")
  ) %>%
  select(
    institution,
    major,
    count,
    debt_median,
    title_iv_count,
    earnings_median,
    earnings_count
  ) %>%
  mutate(
    institution = case_when(
      str_detect(institution, "Olin") ~ "Olin",
      str_detect(institution, "Harvey Mudd") ~ "HMC",
      str_detect(institution, "Rose-Hulman") ~ "RHIT"
    )
  ) %>%
  filter((debt_median > 0) | (earnings_median > 0)) %>%
  arrange(desc(earnings_median)) %>%
  knitr::kable()
```

I know that HMC tends to pride itself on high starting salaries for its
graduates; it appears that CS drives a lot of that outcome. Olin's general
engineering degree includes both CS-focused and other self-designed majors,
which probably accounts for some of the difference with Olin's ME degree.

Given the small numbers at HMC and Olin, it appears the ED does not have access
to data across all majors offered. The ED is probably only able to collect data
from students that interact with federally-funded programs---this is something
to keep in mind when studying the rest of the data.

## EDA
<!-- -------------------------------------------------- -->

What majors tend to receive Title IV funding in greater volume; that is, number
of Title IV counts (not dollar values)?

```{r eda-title-iv}
df_data %>%
  group_by(major) %>%
  summarize(
    total = sum(title_iv_count, na.rm = TRUE),
    mean = mean(title_iv_count, na.rm = TRUE)
  ) %>%
  arrange(desc(total)) %>%
  head(n = 20) %>%
  knitr::kable()
```

- Top Title IV volume goes to Business admin and various health professions

What majors have the greatest debt?

```{r eda-debt}
df_data %>%
  group_by(major) %>%
  summarize(
    total = sum(debt_median, na.rm = TRUE),
    median = median(debt_median, na.rm = TRUE)
  ) %>%
  arrange(desc(median)) %>%
  head(n = 20) %>%
  knitr::kable()
```

Perhaps unsurprisingly, medical doctoral programs result in the greatest debt,
at least within this dataset.

How does total debt compare against earnings for different majors?

```{r earnings-v-debt, fig.width=14, fig.height=10}
df_major_summaries <-
  df_data %>%
  filter(level == 3) %>% # UG only
  group_by(major) %>%
  summarize(
    earnings_median = median(earnings_median, na.rm = TRUE),
    debt_median = median(debt_median, na.rm = TRUE)
  ) %>%
  ## Categorize earnings-debt space
  filter(!is.na(debt_median), !is.na(earnings_median)) %>%
  add_pareto(
    max(debt_median) | max(earnings_median),
    var = "expensive"
  ) %>%
  add_pareto(
    min(debt_median) | max(earnings_median),
    var = "efficient"
  ) %>%
  add_pareto(
    max(debt_median) | min(earnings_median),
    var = "poor"
  ) %>%
  mutate(plain = !expensive & !efficient & !poor) %>%
  gather(key = "type", value = "flag", expensive, efficient, poor, plain) %>%
  filter(flag) %>%
  select(-flag) %>%
  ## Aesthetics
  mutate(major = str_wrap(major, width = 15))

## Plot
df_major_summaries %>%
  filter(type != "plain") %>%
  ggplot(aes(debt_median, earnings_median)) +
  geom_point(
    data = df_major_summaries %>%
      filter(type == "plain") %>%
      select(-type),
    color = "grey80",
    size = 2
  ) +
  geom_point(
    mapping = aes(fill = type),
    color = "black",
    pch = 21,
    size = 2
  ) +
  geom_label_repel(
    mapping = aes(label = major),
    size = 3,
    force = 10
  ) +
  ## Tail
  facet_grid(type~.) +
  scale_x_log10(breaks = c(10000, 30000, 100000)) +
  scale_y_log10(breaks = c(10000, 30000, 100000)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Median Debt",
    y = "Median Earnings",
    title = "Earnings v Debt, Undergraduate Majors",
    caption = "Source: ED College Scorecard Data"
  )
```

I categorize undergraduate degrees based on their location in Earning-Debt
space:
- Efficient majors have high Earnings and low Debt
- Expensive majors have both high Earnings and high Debt
- Poor majors have low Earnings but high Debt

Such information could potentially be helpful to students in major decisions.

## Engineering Major Debt and Earnings
<!-- -------------------------------------------------- -->

I have a particular interest in Engineering degrees; how do these compare?

```{r vis-engr, fig.width=8, fig.height=10}
df_dollars <-
  df_data %>%
  filter(
    str_detect(major, "Engineer"),
    level == 3
  ) %>%
  ## Dollars
  gather(key = "type", value = "val", earnings_median, debt_annual) %>%
  group_by(major, type) %>%
  summarize(val = median(val, na.rm = TRUE)) %>%
  ungroup() %>%
  ## Factor mutation
  mutate(
    major = str_remove(major, "Engineer[[:alpha:]]+\\s?"),
    major = str_remove_all(major, "[,|\\.|-]"),
    major = str_remove(major, "\\s?$"),
    type = case_when(
      type == "debt_annual" ~ "Debt payment",
      type == "earnings_median" ~ "Earnings"
    )
  )

df_counts <-
  df_data %>%
  filter(str_detect(major, "Engineer")) %>%
  ## Dollars
  gather(key = "type", value = "count", title_iv_count, earnings_count) %>%
  mutate(count = if_else(is.na(count), 0, count)) %>%
  group_by(major, type) %>%
  summarize(count = mean(count)) %>%
  ungroup() %>%
  ## Factor mutation
  mutate(
    major = str_remove(major, "Engineer[[:alpha:]]+\\s?"),
    major = str_remove_all(major, "[,|\\.|-]"),
    major = str_remove(major, "\\s?$"),
    type = case_when(
      type == "title_iv_count" ~ "Debt payment",
      type == "earnings_count" ~ "Earnings"
    )
  )

left_join(
  df_dollars,
  df_counts,
  by = c("major", "type")
) %>%
  filter(!is.na(val)) %>%

  ggplot(aes(fct_reorder(major, val), val)) +
  geom_point(
    aes(size = count, fill = type),
    color = "black",
    pch = 21
  ) +
  scale_y_log10(breaks = c(3000, 4000, 40000, 60000, 80000)) +
  scale_fill_discrete(name = "") +
  scale_size_continuous(name = "Mean Count") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Engineering Undergraduate Major",
    y = "Annual",
    title = "Annual Graduate Debt Payment / Earnings",
    caption = "Source: ED College Scorecard Data"
  )
```

Observations:
- Computer Technicians and Electromechnical engineers have outlier levels of high debt
- There are two clear 'winner' degrees in terms of Earnings, and a tail of lower-paying Engineering UG majors
- There is a wide bulk of 'middle'-paying Engineering majors
- Petroleum, chemical, and metallurgical-related majors tend to be on the higher-paid side
- Majors with a 'technician' flavor tend to be less well-paid

A focused version of the same plot, but filtering for a minimum count.

```{r vis-engr-major}
left_join(
  df_dollars,
  df_counts,
  by = c("major", "type")
) %>%
  filter(
    !is.na(val),
    count >= 10
  ) %>%

  ggplot(aes(fct_reorder(major, val), val)) +
  geom_point(
    aes(fill = type),
    color = "black",
    pch = 21,
    size = 4
  ) +
  scale_y_log10(breaks = c(3000, 4000, 40000, 80000)) +
  scale_fill_discrete(name = "") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Engineering Undergraduate Major",
    y = "Annual",
    caption = "Source: ED College Scorecard Data; mean count at least 10",
    title = "Annual Debt Payment / Earnings"
  )
```
