---
title: "Password Data"
author: "Zach"
date: 2020-01-19
output:
  github_document
---

This week's data are [terrible
passwords](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-01-14),
which are all terrible for different reasons. These will be fun!

```{r setup}
library(tidyverse)
library(lubridate)

# Get the Data

df_pass <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-14/passwords.csv') %>%
  filter(!is.na(password))

```

## EDA
<!-- -------------------------------------------------- -->

What are the Top-10 most popular passwords?

```{r head}
df_pass %>% head(10) %>% knitr::kable()
```

A lot of these are (sadly) unsurprising, but I'm surprised to see `dragon` on
here.

Which are the 'best of the worst' passwords?

```{r best-worst}
df_pass %>%
  arrange(desc(offline_crack_sec)) %>%
  head(10) %>%
  knitr::kable()
```

These are really funny. :D

How are passwords distributed among the categories?

```{r categories}
df_pass %>%
  count(category) %>%
  arrange(desc(n))
```

Some of these categories are a bit quizzical; what are "fluffy" passwords?

```{r fluffy}
df_pass %>%
  filter(category == "fluffy")
```

Ah, I see.

What's the distribution of crack-times?

```{r times}
df_pass %>%
  ggplot(aes(offline_crack_sec)) +
  geom_histogram(bins = 5) +

  scale_x_log10(labels = scales::label_number_auto()) +
  theme_minimal() +
  labs(
    x = "Offline Time-to-hack (sec)",
    y = "Count"
  )
```

This seems odd; I wonder what's in that "spike" near 10 milliseconds?

```{r spike-filter}
df_pass %>%
  filter(
    1e-2 / 5 <= offline_crack_sec,
    offline_crack_sec <= 1e-2
  )
```

These passwords seem to be alphabetic and of similar length; let's check the
distribution of lengths against the distribution of hack-times.

```{r length-v-time}
df_pass %>%
  mutate(len = str_length(password)) %>%

  ggplot(aes(len, offline_crack_sec)) +
  geom_count() +

  scale_y_log10() +
  theme_minimal()
```

Yup! We're seeing a lot of passwords of length 6 with similar offline
hack-times. The bands are very interesting structures; there seem to be some
features of passwords (of the same length) that raise or lower their crack-time
by orders of magnitude.

Can we see the aforementioned difference from the given `category` variable?

```{r length-v-time-v-category}
df_pass %>%
  mutate(len = str_length(password)) %>%

  ggplot(aes()) +
  geom_count(
    aes(len, offline_crack_sec, color = category),
    position = position_dodge(width = 0.4)
  ) +
  geom_line(
    data = tibble(len = seq(4, 9)) %>%
      mutate(offline_crack_sec = 1e-10 * 10 ^ len),
    aes(len, offline_crack_sec),
    linetype = 2
  ) +

  scale_y_log10() +
  scale_color_discrete(name = "Category") +
  scale_size_continuous(name = "Count") +
  theme_minimal() +
  labs(
    x = "Password Length",
    y = "Offline Crack-Time (Sec)"
  )
```

Observations:

- Generally, password crack-times grow exponentially with password length. This is why general password advice is "longer is better."
- We can see that the `length==6` spike of `offline_crack_sec ~ 1e-2` can be sub-divided into a large number of categories. Those are:

```{r spike-categories}
df_pass %>%
  filter(
    str_length(password) == 6,
    1e-3 < offline_crack_sec,
    offline_crack_sec <= 1e-2
  ) %>%
  count(category) %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

- The rate of growth of crack-time seems to be greater for the upper band. I suspect the lower band are ordered sequences, which have much more structure.

Let's test that hypothesis! The manual line I drew above separates the bands,
so we can use that to filter.

```{r filter-band}
df_pass %>%
  mutate(len = str_length(password)) %>%
  filter(offline_crack_sec <= 1e-10 * 10 ^ len) %>%
  select(rank, password, category, offline_crack_sec) %>%
  knitr::kable()
```

Ah yes, this pattern makes sense. There is one digit string that's not a
"simple" pattern, but it's definitely in the Zeitgeist.
