---
title: "Daily Data Scrape"
author: "Zach"
date: 2020-03-16
output:
  github_document
---

**Purpose**: Our understanding of the covid-19 outbreak is rapidly developing.
This means our picture of the situation---the data available---are going to
evolve quickly. To build a retrospective picture of how our understanding
evolved, I'm going to track one source of data as it develops, and keep a daily
history.

This document scrapes the Santa Clara County government site for covid-19
updates, and saves each day's version.

```{r setup}
library(tidyverse)
library(rvest)
library(lubridate)

filename_stub <- "./data/covid19_santa_clara"
date_today <- today()
```

## Load and Wrangle
--------------------------------------------------

Load the Santa Clara update page

```{r santa-clara}
url_santa_clara <- "https://www.sccgov.org/sites/phd/DiseaseInformation/novel-coronavirus/Pages/home.aspx"
css_santa_clara <- "#MSOZoneCell_WebPartctl00_m_g_6ee5292c_2758_48a4_a561_9decdb5a0be9 > table:nth-child(1)"
css_santa_header <- "h2.ms-rteElement-H2:nth-child(4)"
page <- read_html(url_santa_clara)
```

Scrape the update table

```{r scrape}
str_santa_clara <-
  page %>%
  html_nodes(css_santa_clara) %>%
  html_table() %>%
  .[[1]] %>%
  .[[1]]

## Doesn't extract cleanly; grab the string
str_santa_clara
```

Wrangle the table

```{r get-data}
## Split the string and look for colon separators
df_today <-
  str_split(
    str_santa_clara,
    ","
  ) %>%
  as_tibble(.name_repair = "unique") %>%
  rename(s = `...1`) %>%
  mutate(s = str_remove_all(s, "\"")) %>%
  extract(s, into = c("var", "val"), regex = "(\\w+):(\\d+)", convert = TRUE) %>%
  filter(!is.na(val))
```

Parse the header for the update time.

```{r get-update-date}
date_update <-
  page %>%
  html_nodes(css_santa_header) %>%
  html_text() %>%
  str_extract(., "as of .*") %>%
  str_extract(., "[:upper:].*$") %>%
  mdy(.)
```

## Write
<!-- -------------------------------------------------- -->

```{r write}
filename <- str_c(
  filename_stub,
  "_t",
  date_today,
  "_u",
  date_update,
  ".csv"
)

write_csv(df_today, filename)
```
