---
title: "Measles Vaccines"
author: "Zach"
date: 2020-02-25
output:
  github_document
---

```{r setup}
library(tidyverse)
library(sf)

measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-25/measles.csv')

theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin   = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title    = element_text(size = 18, hjust = 0),
    plot.subtitle = element_text(size = 16),
    plot.caption  = element_text(size = 14, hjust = 1)
  )
}
```

```{r dl-boundaries}
url_boundaries <- "https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_state_20m.zip"
file_boundaries = "../data/usa_states.zip"

curl::curl_download(url_boundaries, destfile = file_boundaries)
unzip(file_boundaries, exdir = "../data")

sf_boundaries <- sf::read_sf("../data", "cb_2018_us_state_20m")
```

### Data dictionary

|variable |class     |description |
|:--------|:---------|:-----------|
|index    |double    | Index ID |
|state    |character | School's state |
|year     |character | School's district|
|name     |character | School name|
|type     |character | Whether a school is public, private, charter |
|city     |character | City |
|county   |character | County |
|district |logical   | School district |
|enroll   |double    | Enrollment |
|mmr      |double    | School's Measles, Mumps, and Rubella (MMR) vaccination rate |
|overall  |double    | School's overall vaccination rate|
|xrel     |logical   | Percentage of students exempted from vaccination for religious reasons |
|xmed     |double    | Percentage of students exempted from vaccination for medical reasons |
|xper     |double    | Percentage of students exempted from vaccination for personal reasons |

```{r summary}
measles %>% summary
```

Oddly, only a subset of the states are listed:

```{r }
measles %>% pull(state) %>% unique
measles %>% pull(state) %>% unique %>% length
```

We only have `32` of the states in the dataset.

## Post-processing
<!-- -------------------------------------------------- -->

There's a bunch of odd stuff in these data; below I document some observations
and wrap up with some post-processing before analysis.

### Variable `year`

Year is a character, and the data dictionary calls it the "school's district".
That seems odd...

```{r year-unique}
measles %>% pull(year) %>% unique
```

Assume that the `2017` level is the beginning of the academic year; then we can
pull the first four digits to get the year.

### Missing data coding

It appears that `mmr` and `overall` use the value `-1` to encode missing values.
However, this will mess with any aggregation we do later. Let's replace
out-of-bounds values with `NA`.

### Post-processing

Run post-processing to fix the issues found above.

```{r proc}
df_measles <-
  measles %>%
  mutate(
    ## Variable `year`
    year = str_extract(year, "^\\d{4}") %>%
      as.integer,
    ## Missing data coding
    mmr = if_else(mmr < 0, NA_real_, mmr),
    overall = if_else(overall < 0, NA_real_, overall)
  )
```

## EDA
<!-- -------------------------------------------------- -->

```{r }
df_measles %>%
  select(state, mmr, overall) %>%
  pivot_longer(
    mmr:overall,
    values_to = "percent",
    names_to = "type"
  ) %>%
  group_by(state, type) %>%
  mutate(
    percent_mu = median(percent, na.rm = TRUE),
    percent_sd = sd(percent, na.rm = TRUE),
  ) %>%
  ungroup() %>%
  mutate(state = fct_reorder(state, percent_mu)) %>%

  ggplot(aes(state, percent, fill = type)) +
  geom_boxplot(
    position = position_dodge(width = 1)
  ) +
  coord_flip()
```

Observations:
-

```{r}
sf_measles <-
  st_as_sf(
    df_measles %>% filter(!is.na(lng), !is.na(lat)),
    coords = c("lng", "lat"),
    crs = 4326,
    agr = "constant"
  )
sf_measles %>% glimpse
```

```{r}
left_join(
  sf_boundaries,
  df_measles %>%
    group_by(state) %>%
    summarize(mmr = median(mmr, na.rm = TRUE)),
  by = c("NAME" = "state")
) %>%
  filter(!(NAME %in% c("Alaska", "Hawaii", "Puerto Rico"))) %>%

  ggplot() +
  geom_sf(aes(fill = mmr)) +

  viridis::scale_fill_viridis(name = "Rate (%)") +
  guides(
    fill = guide_colorbar(barwidth = 10)
  ) +
  theme_common() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    aspect.ratio = 1 / 2,
    legend.position = "bottom"
  ) +
  labs(
    title = "Statewide Median MMR Vaccination Rate",
    caption = "Gray state data unavailable"
  )
```
