---
title: "Philadelphia Parking Violations"
author: "Zach"
date: 2019-12-03
output:
  github_document
---

```{r setup}
library(tidyverse)
library(lubridate)
library(modelr)
```

```{r load, cache=TRUE}
df_tickets <-
  readr::read_csv(
"https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-03/tickets.csv"
)
```

## Simple Checks
--------------------------------------------------

```{r glimpse}
df_tickets %>% glimpse
```

- There are `1,260,891` tickets in this dataset.

```{r summary}
df_tickets %>% summary
```

- The records are for 2017, spanning January 1st until December 31st
- Fines are mostly for small amounts of cash, though the most expensive ticket is for over a thousand dollars!
- Zip codes are given for nearly all tickets, though a small minority (in the 100,000's) are not available

```{r desc-levels}
descriptions <-
  df_tickets %>%
  pull(violation_desc) %>%
  unique

length(descriptions)
descriptions
```

- There are `95` ticket descriptions

```{r agency-levels}
df_tickets %>%
  pull(issuing_agency) %>%
  unique
```

I hadn't thought about it before, but it stands to reason that multiple agencies
have the authority to issue tickets. The `POLICE` are of course here, but `PENN`
presumably refers to Pennsylvania University.

## 1 Dimensional EDA
<!-- -------------------------------------------------- -->

I tried a histogram of the `fine` values, but the fingers were ridiculous.
Checking a count table shows us why:

```{r count-fines}
df_tickets %>%
  count(fine)
```

- Fines seem to come in standard values

```{r vis-fines}
df_tickets %>%
  count(fine) %>%

  ggplot(aes(fine, n)) +
  geom_bar(stat = "identity") +
  ## Tail
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    x = "Fine ($)",
    y = "Count"
  )
```

Observations:

- The bulk of tickets are in the < $100 range
- A minority of tickets are in the $100-$300 range
- A small fraction of tickets are around $1000
  + What violation *is* this?!

## Timeseries
<!-- -------------------------------------------------- -->

Let's try

```{r date-issue, fig.width=12}
df_tickets %>%
  mutate(issue_date = date(issue_datetime)) %>%
  count(issue_date) %>%

  ggplot(aes(issue_date, n)) +
  geom_bar(stat = "identity") +
  ## Tail
  theme_minimal() +
  theme(aspect.ratio = 1 / 3) +
  labs(
    x = "Date of Issue",
    y = "Count"
  )
```

Observations:

- There seems to be some cyclical pattern in the data; this is probably the effect of a work-week
- It's rather challenging to see much from this visualization; let's try modeling the week
effect and subtracting it.

Here I model the weekday effect on the counts.

```{r weekday-issue}
df_dates <-
  df_tickets %>%
  mutate(issue_date = date(issue_datetime)) %>%
  count(issue_date) %>%
  mutate(issue_wday = wday(issue_date, label = TRUE))

fit_wday <-
  lm(
    formula = n ~ issue_wday,
    data = df_dates
  )

df_dates <-
  df_dates %>%
  add_predictions(fit_wday, var = "n_hat") %>%
  mutate(n_resid = n - n_hat)
```

Let's visualize the weekday-effect model:

```{r counts-weekdays}
tibble(
  issue_wday = df_dates %>%
    pull(issue_wday) %>%
    unique
) %>%
  add_predictions(fit_wday, var = "n_hat") %>%

  ggplot(aes(issue_wday, n_hat)) +
  geom_bar(stat = "identity") +
  theme_minimal()
```

The peak issue-date seems to be Thursday, with Sunday having the fewest tickets issued.
The low weekend counts can explain the cyclical pattern we saw above.

```{r resid, fig.width=12}
df_holidays <-
  tribble(
           ~date, ~holiday,
    "2017-01-01", "New Year",
    "2017-01-16", "MLK Day",
    "2017-02-14", "Valentine",
    "2017-05-29", "Memorial",
    "2017-07-04", "Indep. Day",
    "2017-09-04", "Labor Day",
    "2017-11-10", "Veterans Day",
    "2017-11-23", "Thanksgiving",
    "2017-12-25", "Christmas"
  ) %>%
  mutate(date = ymd(date))

df_dates %>%
  ggplot(aes(issue_date, n_resid)) +
  geom_bar(stat = "identity") +
  geom_segment(
    data = df_holidays,
    mapping = aes(x = date, y = 0, xend = date, yend = 900),
    color = "red",
    size = 0.2
  ) +
  geom_text(
    data = df_holidays,
    mapping = aes(x = date, y = 1000, label = holiday),
    angle = 90,
    hjust = 0
  ) +
  ## Tail
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(limits = c(-4000, 2500)) +
  theme_minimal() +
  theme(aspect.ratio = 1 / 3) +
  labs(
    x = "Issue Date",
    y = "Seasonal Effect",
    title = "Seasonal Effect on Number of Parking Violations",
    subtitle = "Philadelphia in 2017"
  )
```

Negative spikes correspond to fewer tickets than weekday effects would predict

There do seem to be **strong** seasonal effects:

- Many of the strong negative spikes are associated with national holidays.
- The time around Christmas day seems to be a period of prolonged low-ticket behavior; sensibly, people are probably staying home during this time.
- March 14th is an oddity; the best I can tell is there was a [heavy snowstorm](https://www.inquirer.com/philly/blogs/real-time/Snow-sleet-freezing-rain-Stella-blizzard-Philadelphia-March-14-2017.html) that week.
- February 9th may also be explained by a [snowstorm](https://www.phillyvoice.com/latest-forecast-goes-deep-6-10-inches-of-snow-across-region/).
