---
title: "ABET-Accredited Engineering Programs in the US"
author: "Zach"
date: 2020-04-09
output:
  github_document
---

A simple investigation of Wikipedia's [List of engineering
schools](https://en.wikipedia.org/wiki/List_of_engineering_schools). My focus is
on institutions in the US.

```{r setup}
library(tidyverse)
library(rvest)

url_engr <- "https://en.wikipedia.org/wiki/List_of_engineering_schools"
engr_selector <- "table.wikitable:nth-child(42)"
```

## Scrape and Wrangle Data
<!-- -------------------------------------------------- -->

```{r }
page <- read_html(url_engr)
```

```{r }
## Parse html
df_raw <-
  page %>%
  html_nodes(engr_selector) %>%
  html_table(fill = TRUE) %>%
  .[[1]] %>%
  as_tibble()

## Wrangle
df_engr <-
  df_raw %>%
  mutate(
    `State (city)[10]` = if_else(
      str_detect(`State (city)[10]`, "District of Columbia"),
      "District of Columbia (District of Columbia)",
      `State (city)[10]`
    )
  ) %>%
  extract(
    col = `State (city)[10]`,
    into = c("state", "city"),
    regex = "([\\w\\s[:punct:]]+)\\((.+)\\)"
  ) %>%
  rename(
    has_grad = `Master's/doctoralprograms?`
  ) %>%
  rename_all(str_to_lower) %>%
  mutate(
    has_grad = if_else(has_grad == "Yes", TRUE, FALSE),
    state = str_trim(state)
  )

df_engr %>%
  filter(is.na(state))
```

## EDA
<!-- -------------------------------------------------- -->

```{r }
df_engr %>% count()
```

There are `r df_engr %>% count() %>% pull(n)` universities in the United States
with ABET-accredited engineering programs. This is fewer than I expected.

```{r }
df_engr %>%
  count(state) %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

Observations:

- California, Texas, and New York have the most institutions with engineering programs
- Delaware, Hawaii, Nebraska, and Wyoming have the fewest (one each)

```{r }
df_engr %>%
  count(city) %>%
  arrange(desc(n)) %>%
  head(10)
```

Observations:

- The cities with the most engineering institutions are the ones I'd expect
- However, I'd think Boston would have more institutions?
  - Note that MIT and Harvard are in Cambridge
