---
title: "San Francisco Trees"
author: "Zach"
date: 2020-01-28
output:
  github_document
---

## Setup
<!-- ----------------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
library(sf)

df_trees <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-28/sf_trees.csv')

```

```{r}
sf_trees <- 
  st_as_sf(
    df_trees %>% filter(!is.na(latitude), !is.na(longitude)),
    coords = c("longitude", "latitude"),
    crs = 4326,
    agr = "constant"
  )
sf_trees %>% glimpse
```

```{r}
species_many <- 
  sf_trees %>% 
  count(species) %>% 
  filter(n > 100) %>% 
  pull(species)

species_few <- 
  sf_trees %>% 
  count(species) %>% 
  filter(n < 100) %>% 
  pull(species)
  
```


### Download

```{r get-parcels}
url_parcels <- "https://data.sfgov.org/api/geospatial/us3s-fp9q?method=export&format=Shapefile"
curl::curl_download(url_parcels, destfile = "../data/parcels.zip")
unzip("../data/parcels.zip", exdir = "../data")
## Record filename
names_parcels <- unzip("../data/parcels.zip", list = TRUE)
main_parcels <- 
  names_parcels %>% 
  pull(Name) %>% 
  .[[1]] %>% 
  str_remove(., "\\.\\w+$")

sf_parcels <- read_sf(
  "../data", 
  main_parcels
)
```

```{r get-streets}
url_streets <- "https://data.sfgov.org/api/geospatial/3psu-pn9h?method=export&format=Shapefile"
curl::curl_download(url_streets, destfile = "../data/streets.zip")
unzip("../data/streets.zip", exdir = "../data")
## Record filename
names_streets <- unzip("../data/streets.zip", list = TRUE)
main_streets <- 
  names_streets %>% 
  pull(Name) %>% 
  .[[1]] %>% 
  str_remove(., "\\.\\w+$")

sf_streets <- read_sf(
  "../data", 
  main_streets
)
```

```{r}
sf_streets %>% glimpse
```


## Vis
<!-- ----------------------------------------------------------------------------- -->

Let's go looking for *special* trees.

```{r special}
ggplot() +
  geom_sf(
      data = sf_streets %>% 
        mutate(
          layer = str_to_lower(layer),
          highway = str_detect(layer, "freeway")
        ),
      mapping = aes(size = highway)
    ) +
  geom_sf(
    data = sf_trees %>% filter(species %in% species_few),
    mapping = aes(color = species),
    size = 0.2
  ) +
  
  scale_size_manual(values = c(`TRUE` = 0.1, `FALSE` = 0.05)) +
  guides(size = FALSE, color = FALSE) +
  theme_minimal()
```

Observations:

- I see a few "lines" of special trees of the same species

Let's zoom in to some of these special rows:

```{r, fig.width=12, fig.height=8}
bound = st_polygon(
    list(rbind(
      c(-122.447, +37.778), 
      c(-122.447, +37.780), 
      c(-122.440, +37.780), 
      c(-122.440, +37.778), 
      c(-122.447, +37.778)
    ))
  )

sf_streets %>% 
  filter(
    st_within(., bound, sparse = FALSE) | 
    st_intersects(., bound, sparse = FALSE)
  ) %>% 
  mutate(
    layer = str_to_lower(layer),
    highway = str_detect(layer, "freeway"),
    center_lon = map_dbl(geometry, ~st_centroid(.)[[1]]),
    center_lat = map_dbl(geometry, ~st_centroid(.)[[2]])
  ) %>% 
ggplot() +
  geom_sf(aes(size = highway)) +
  geom_label(
    data = . %>% 
      group_by(streetname) %>% 
      filter(min_rank(center_lon) <= 1) %>% 
      ungroup(),
    aes(center_lon, center_lat, label = str_to_title(streetname))
  ) +
  geom_sf(
    data = sf_trees %>% 
      filter(
        st_within(., bound, sparse = FALSE),
        species %in% species_few
      ),
    mapping = aes(color = species),
    size = 1.0
  ) +
  
  scale_size_manual(values = c(`TRUE` = 0.1, `FALSE` = 0.05)) +
  guides(size = FALSE) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Longitude",
    y = "Latitude"
  )

```

There's (apparently) a long row of `American Ash` trees on Turk Blvd. near Arbol Lane. This is a strange place for American Ash, whose natural range is the [Eastern US](https://en.wikipedia.org/wiki/Fraxinus_americana).

Based on a [streetview](https://goo.gl/maps/iYQHZJ8zG61HqNoF8), it looks like Arbol Lane is actually a staircase.

