---
title: "Christmas Songs on the Hot-100"
author: "Zach"
date: 2019-12-25
output:
  github_document
---

Tidy Tuesday 2019-12-24 dataset: Christmas songs on the Top-100 list.

```{r setup}
library(tidyverse)
library(tidytext)
library(lubridate)

df_songs <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-24/christmas_songs.csv") %>%
  mutate(time = mdy(weekid))

df_lyrics <- readr::read_tsv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-24/christmas_lyrics.tsv")

```

## Glimpses
<!-- -------------------------------------------------- -->

A quick look at the data:

```{r glimpse-songs}
df_songs %>% glimpse
```

```{r glimpse-lyrics}
df_lyrics %>% glimpse
```

I'm not sure why there are both `title` and `track_title` variables...

```{r lyrics-title}
df_lyrics %>% pull(title) %>% unique
```

There are only `6` unique `title` values.

```{r lyrics-track_title}
df_lyrics %>% pull(track_title) %>% unique
```

There are `49` unique `track_title` values; this is what I'll use to group the
song lyrics.

```{r lyrics-len}
df_lyrics %>%
  group_by(track_title) %>%
  summarize(length = max(line)) %>%
  select(track_title, length)
```

We don't have lyric data on every song; let's see how many have this info.

```{r lyrics-len-info}
df_lyrics %>%
  group_by(track_title) %>%
  summarize(length = max(line)) %>%
  select(track_title, length) %>%
  filter(!is.na(length)) %>%
  arrange(desc(length))
```

There are `35` tracks (out of `49`) with lyric data in this set. I'll filter
down to just those songs for which we have lyrics.

## Timeseries
<!-- -------------------------------------------------- -->

First, let's try a basic visualization of all the Top-100 Christmas songs:

```{r songs-timeseries-vis}
df_songs %>%
  mutate(title = str_to_title(song)) %>%

  ggplot(
    aes(time, week_position, color = fct_reorder2(title, time, week_position))
  ) +
  geom_step(
    aes(),
    direction = "hv"
  ) +
  geom_point() +
  ## Tail
  guides(color = FALSE) +
  theme_minimal()
```

From this vis we can tell a few things:

- Christmas songs on the Top-100 are fairly rare; there was a large collection
  of hits in the 1960's, a short run in the 1970's, and a resurgance only in the
  past decade.
- A few songs have enjoyed a "revival", as evidenced by the long horizontal
  lines connecting the same song charting decades across.

What are these "revived" Christmas songs?

```{r songs-revived}
df_songs %>%
  group_by(song) %>%
  mutate(time_delta = max(time) - min(time)) %>%
  mutate(
    case = case_when(
      percent_rank(time) == 0 ~ "Earliest",
      percent_rank(time) == 1 ~ "Latest",
      TRUE ~ NA_character_
    ),
    time_perf = str_c(time, " ", performer)
  ) %>%
  ungroup() %>%
  arrange(desc(time_delta)) %>%
  filter(!is.na(case), time_delta > years(5)) %>%
  select(song, time_perf, case) %>%
  pivot_wider(names_from = case, values_from = time_perf)
```

There are been `7` "revived" Christmas songs, where the earliest to latest
charting was over 5 years apart. Most of these are attributed to the same
artist, though a few have been re-interpreted by a different performer. The
longest span is (shudder) "Jingle Bell Rock", which charted as recently as 2017.

## Process Lyrics
<!-- -------------------------------------------------- -->

Let's tokenize the lyrics for each song (split into words) and join sentiment
data for each entry. Since I'm using the `afinn` numeric scale, I zero-fill
those words without sentiment.

```{r lyrics-tokenize, warning=FALSE}
df_lyrics_tokens <-
  df_lyrics %>%
  select(track_title, line, lyric) %>%
  filter(!is.na(line)) %>%
  unnest_tokens(word, lyric) %>%
  left_join(
    .,
    get_sentiments("bing"),
    by = "word"
  ) %>%
  group_by(track_title) %>%
  mutate(
    song_n = row_number(),
    song_p = song_n / max(song_n),
    sentiment = if_else(is.na(sentiment), "neutral", sentiment),
    sentiment = fct_relevel(sentiment, "negative", "neutral", "positive"),
    value = case_when(
      sentiment == "negative" ~ -1,
      sentiment == "neutral" ~ 0,
      sentiment == "positive" ~ +1,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()

df_lyrics_tokens %>% glimpse
```

How many unique words show up in each song?

```{r lyrics-count-unique-words}
df_unique_words <-
  df_lyrics_tokens %>%
  select(track_title, word) %>%
  unique() %>%
  count(track_title) %>%
  arrange(desc(n))

df_unique_words %>% head
df_unique_words %>% tail
```

The songs `r df_unique_words %>% head %>% pull(track_title)` have the most
words, while `r df_unique_words %>% tail %>% pull(track_title)` have the fewest
words.

Let's count the fraction of unique words in each song. The lower the fraction,
the more repetitive the lyrics.

```{r lyrics-count-unique-frac}
df_unique_frac <-
  df_lyrics_tokens %>%
  select(track_title, word) %>%
  group_by(track_title) %>%
  mutate(n_tot = n()) %>%
  unique() %>%
  mutate(
    n_words = n(),
    frac = n_words / n_tot
  ) %>%
  ungroup() %>%
  select(track_title, frac) %>%
  unique() %>%
  arrange(desc(frac))

df_unique_frac %>% head
df_unique_frac %>% tail
```

Observations:

- "The Christmas Song" is the least repetitive song
- "Jingle Bells" is the most repetitive, followed closely by "Jingle Bell Rock"

```{r lyrics-frac-vis}
df_unique_frac %>%
  ggplot(aes(frac)) +
  geom_histogram(bins = 35) +
  theme_minimal()
```

Observations:

- It's not just a feeling: Christmas songs tend to be very repetitive....

```{r lyrics-frac-mean}
df_unique_frac %>%
  summarize(frac_mean = mean(frac))
```

In the songs for which we have data, the lyrics tend to be very repetitive. On
average, if the lyrics have 100 words, the song itself will tend to repeat the
same 18 words over and over.

Is there any relation between max Top-100 rank and repetitiveness?

```{r rank-v-frac}
p <-

  left_join(
    df_unique_frac %>% mutate(title = str_to_lower(track_title)),
    df_songs %>% mutate(title = str_to_lower(song)),
    by = "title"
  ) %>%
  select(title, frac, week_position) %>%
  mutate(week_position = if_else(is.na(week_position), 0, week_position)) %>%
  group_by(title) %>%
  filter(min_rank(-week_position) == 1) %>%
  ungroup() %>%
  mutate(title = str_to_title(title)) %>%

  ggplot(aes(frac, week_position)) +
  geom_vline(xintercept = 0, size = 2, color = "grey80") +
  ggrepel::geom_label_repel(
    data = . %>% filter(week_position >= 1),
    aes(label = str_wrap(title, width = 20), fill = title),
    point.padding = 1,
    size = 2,
    hjust = 0,
    seed = 102
  ) +
  geom_point(
    data = . %>% filter(week_position < 1),
    color = "grey60",
    position = position_jitter(width = 0, height = 0.5, seed = 101)
  ) +
  geom_point(
    data = . %>% filter(week_position >= 1),
    aes(color = title)
  ) +
  ## Tail
  scale_y_continuous(limits = c(0, 110)) +
  guides(color = FALSE, fill = FALSE) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Unique Word Fraction",
    y = "Max Top-100 Rank"
  )

ggExtra::ggMarginal(p, type = "histogram", margins = "x")
```

There are not many observations where we have both Top-100 rankings and lyrics
data, so we probably can't draw many conclusions from this. It's interesting to
note that not many of the high-fraction songs made it into the Top-100, but we
can't support that conclusion entirely due to the missing data.

## Word Sentiment Analysis
<!-- -------------------------------------------------- -->

I decided to use this as an excuse to learn how to do simple [sentiment
analysis](https://www.tidytextmining.com/sentiment.html). For each song, how
many unique words with non-zero sentiment are there?

```{r lyrics-nz-words}
df_unique_value_words <-
  df_lyrics_tokens %>%
  select(track_title, word, value) %>%
  unique() %>%
  filter(value != 0) %>%
  count(track_title) %>%
  arrange(desc(n))

df_unique_value_words %>% head
df_unique_value_words %>% tail
```

A couple issues with this analysis:

- The sentiment database I use is English only; this won't work well for some
  of the Spanish songs, e.g.

```{r lyrics-las-posadas}
df_lyrics_tokens %>% filter(track_title == "Las Posadas")
```

- This may be a misleading way to judge "degree sentimentality"; this isn't
  normalizing by the total count of unique words.

Let's instead compute the fraction of unique words which have non-zero
sentiment:

```{r lyrics-frac-words}
df_lyrics_frac <-
  df_lyrics_tokens %>%
  select(track_title, word, value) %>%
  unique() %>%
  group_by(track_title) %>%
  mutate(
    nz = (value != 0)
  ) %>%
  summarize(
    n_nz = sum(nz),
    n_tot = n()
  ) %>%
  mutate(frac = n_nz / n_tot) %>%
  arrange(desc(frac))

df_lyrics_frac %>% head
df_lyrics_frac %>% tail
```

Observations:

- We're seeing similar songs at the bottom: `Las Posadas` shows up again, likely
  due to the language issue.
- Some classic Christmas songs show up in the high-fraction range.

Let's compute the mean sentiment for each song:

```{r lyrics-mean-sentiment}
df_stat_sentiment <-
  df_lyrics_tokens %>%
  group_by(track_title) %>%
  summarize(
    mean_value = mean(value),
    sd_value = sd(value)
  )

df_stat_sentiment %>% arrange(desc(mean_value)) %>% head
df_stat_sentiment %>% arrange(desc(mean_value)) %>% tail
```

Unsurprisingly, Christmas songs tend to use positive-sentiment words. Even the
lowest-sentiment Christmas songs (in this dataset) are only marginally negative.

Which songs have the greatest variability in the sentiment of their words?

```{r lyrics-sd-sentiment}
df_stat_sentiment %>% arrange(desc(sd_value)) %>% head
df_stat_sentiment %>% arrange(desc(sd_value)) %>% tail
```

## Sentiment Visualizations
<!-- -------------------------------------------------- -->

Is there any relation with Hot-100 max-position?

```{r rank-v-sentiment}
p <-

  left_join(
    df_stat_sentiment %>% mutate(title = str_to_lower(track_title)),
    df_songs %>% mutate(title = str_to_lower(song)),
    by = "title"
  ) %>%
  select(title, mean_value, sd_value, weekid, week_position) %>%
  filter(!str_detect(title, "feliz navidad")) %>%
  mutate(week_position = if_else(is.na(week_position), 0, week_position)) %>%
  group_by(title) %>%
  filter(min_rank(-week_position) == 1) %>%
  ungroup() %>%
  mutate(title = str_to_title(title)) %>%

  ggplot(aes(mean_value, week_position)) +
  geom_vline(xintercept = 0, size = 2, color = "grey80") +
  ggrepel::geom_label_repel(
    data = . %>% filter(week_position >= 1),
    aes(label = str_wrap(title, width = 20), fill = title),
    point.padding = 1,
    size = 2,
    hjust = 0,
    seed = 102
  ) +
  geom_point(
    data = . %>% filter(week_position < 1),
    color = "grey60",
    position = position_jitter(width = 0, height = 0.5, seed = 101)
  ) +
  geom_point(
    data = . %>% filter(week_position >= 1),
    aes(color = title)
  ) +
  ## Tail
  guides(color = FALSE, fill = FALSE) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Mean Sentiment",
    y = "Max Top-100 Rank"
  )

ggExtra::ggMarginal(p, type = "histogram", margins = "x")
```

There's no obvious relation between mean sentiment and max Top-100 rank. If
anything, this shows that the ranking songs tend to be in the middle of the
distribution of Christmas songs, in terms of mean sentiment. Note however that
this only considers those songs for which we have lyrics; some of the Top-100
ranked songs lie outside our lyrics database.

Visualize the timeseries of word sentiments.

```{r lyrics-vis-top}
vis_value_timeseries <- function(df) {
  df %>%
    ggplot(aes(song_p, value)) +
    geom_line(size = 0.1, color = "grey80") +
    geom_point(
      data = . %>% filter(value != 0),
      aes(color = track_title),
      size = 0.5
    ) +
    geom_point(
      data = . %>% filter(value == 0),
      color = "grey80",
      size = 0.1
    ) +
    ## Tail
    scale_color_discrete(name = "") +
    scale_x_continuous(labels = scales::label_percent()) +
    facet_wrap(~track_title) +
    guides(color = FALSE) +
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(
      x = "Song Position",
      y = "Word Sentiment"
    )
}
```

High-mean lyrics

```{r vis-mean-head}
sel_track_title <-
  df_stat_sentiment %>%
  arrange(desc(mean_value)) %>%
  head %>%
  pull(track_title)

df_lyrics_tokens %>%
  filter(track_title %in% sel_track_title) %>%
  vis_value_timeseries()
```

High-variability lyrics

```{r vis-sd-head}
sel_track_title <-
  df_stat_sentiment %>%
  arrange(desc(sd_value)) %>%
  head %>%
  pull(track_title)

df_lyrics_tokens %>%
  filter(track_title %in% sel_track_title) %>%
  vis_value_timeseries()
```

Observations:

- The high-variability lyrics are due to liberal use of positive- and
  negative-sentiment words.

Low-variability lyrics

```{r vis-tail}
track_title_tail <-
  df_stat_sentiment %>%
  arrange(desc(sd_value)) %>%
  tail %>%
  pull(track_title)

df_lyrics_tokens %>%
  filter(track_title %in% track_title_tail) %>%
  vis_value_timeseries()
```

Observations:

- The low-variability lyrics are due mostly to songs with neutral charged words.
