---
title: "UNDOC Homicide Data"
author: "Zach"
date: 2019-12-10
output:
  github_document
---

This week's
[tidytuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-12-10)
is focused on re-creating plots from the news. However, I was more interested in
apparent discrepancies between the datasets, and decided to return to the
original UNDOC data.

```{r setup}
library(tidyverse)
library(readxl)

murders <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-10/international_murders.csv")

gun_murders <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-10/gun_murders.csv")

diseases <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-10/diseases.csv")

nyc_regents <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-12-10/nyc_regents.csv")

theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,
    plot.margin   = unit(c(t = +0, b = +0, r = +0, l = +0), "cm")
  )
}
```

Let's look at the two death-related datasets:

```{r death-cf}
murders %>% arrange(desc(count)) %>% head()
gun_murders %>% arrange(desc(count)) %>% head()
```

Both datasets are given as counts per 100,000 persons. The values are
inconsistent---they're probably from different years. I'm curious about the
trends over time, and the relation of these firearm homicide rates with the
general homicide rates. To study that, I'll need to go back to the original data.

## Comparison: UNDOC Homicide Data
<!-- -------------------------------------------------- -->

Data from [UNDOC Global Study on Homicide](https://www.unodc.org/gsh/en/data.html).

```{r undoc-import}
df_raw <- read_excel(
  "GSH2013_Homicide_count_and_rate.xlsx",
  sheet = 1,
  skip = 6,
  col_names = c(
    "region",
    "sub_region",
    "territory",
    "source",
    "org",
    "indicator",
    "2000",
    "2001",
    "2002",
    "2003",
    "2004",
    "2005",
    "2006",
    "2007",
    "2008",
    "2009",
    "2010",
    "2011",
    "2012"
  )
)
```

```{r undoc-wrangle}
## Trim head and notes
df_tmp <-
  df_raw %>%
  slice(-1) %>%
  slice(-(n()-5):-n())

## Detect notes and fill with NA
df_tmp %>%
  mutate(region = if_else(str_detect(region, "estimate|data"), NA_character_, region))

## Repeatedly lag-fill until NA's are gone
countna <- function(df) {
  df %>%
    filter_at(vars(region, sub_region, territory, source, org), any_vars(is.na(.))) %>%
    dim %>%
    .[[1]]
}

while (countna(df_tmp) > 0) {
  df_tmp <-
    df_tmp %>%
    mutate_at(vars(region, sub_region, territory, source, org), ~if_else(is.na(.), lag(.), .))
}

## Cast all the estimates
df_tmp <-
  df_tmp %>%
  mutate_at(
    vars(`2000`:`2012`),
    as.numeric
  )

## Reshape
df_tmp <-
  df_tmp %>%
  pivot_longer(
    `2000`:`2012`,
    names_to = "year",
    values_to = "value",
    names_ptypes = list(year = integer())
  ) %>%
  pivot_wider(
    names_from = indicator,
    values_from = "value"
  ) %>%
  glimpse

## Final cleaning
df_undoc <-
  df_tmp %>%
  rename_all(str_to_lower)

df_undoc %>%
  glimpse
```

Let's check some summaries:

```{r undoc-summary}
df_undoc %>% pull(region) %>% unique
df_undoc %>% summary
```

Sadly, this spreadsheet seems to be about half empty.

Before joining with the firearm homicide data, let's do some quick EDA on the
general homicide data.

```{r undoc-region-timeseries}
df_undoc %>%
  ggplot(aes(year, rate)) +
  geom_smooth(aes(color = fct_reorder2(region, year, rate)), se = FALSE) +
  ## Tail
  scale_y_log10() +
  scale_color_discrete(name = "Region") +
  theme_minimal()
```

Observations:

- I'm surprised to see the Americas have the highest homicide rates! I wonder
  which specific regions are affected?
- Africa has a fairly unstable homicide rate; it has fluctuated quite a bit in
  the early part of the millennium.

```{r undoc-sub_region-timeseries}
df_undoc %>%
  ggplot(aes(year, rate)) +
  geom_smooth(aes(color = fct_reorder2(sub_region, year, rate)), se = FALSE) +
  ## Tail
  scale_y_log10() +
  scale_x_continuous(breaks = c(2000, 2003, 2006, 2009, 2012)) +
  scale_color_discrete(name = "Region") +
  theme_minimal()
```

Observations:

- South and Central America account for the high average homicide rate we saw above.
- Alarmingly Middle Africa has a sharp uptick in homicidate rate starting around 2010.

```{r }
df_undoc %>%
  filter(sub_region == "Northern America") %>%
  group_by(territory) %>%
  summarize(rate = mean(rate, na.rm = TRUE))
```

Observations:

- Greenland has an enormous homicide rate, but keep in mind their population is
  about 56,000 total (so the statistic itself can fluctuate widely).
- The USA has a homicide rate over three times that of neighboring Canada.

```{r vis-general-homicides}
territories <-
  c(
    "United States of America",
    "Canada",
    "Germany",
    "Italy",
    "Spain"
  ) %>%
  paste(., collapse = "|")

df_undoc %>%
  filter(str_detect(territory, territories)) %>%

  ggplot(aes(year, rate)) +
  geom_smooth(aes(color = fct_reorder2(territory, year, rate)), se = FALSE) +
  ## Tail
  scale_y_log10() +
  scale_x_continuous(breaks = c(2000, 2003, 2006, 2009, 2012)) +
  scale_color_discrete(name = "Country") +
  theme_common() +
  labs(
    x = "Year",
    y = "Homicides per 100,000 Persons",
    caption = "Source: UNDOC Homicide Statistics"
  )
```

This gives us a sense of all homicides, but we'll need firearm data to
understand questions about gun violence.

## Join: UNDOC Firearm data
<!-- -------------------------------------------------- -->

I can't find a direct link on the UNDOC website; I found the following
spreadsheet via Google search.

```{r gun-import}
df_gun_raw <-
  read_excel(
    "Homicides_by_firearms.xls",
    sheet = 1,
    skip = 6,
    col_names = c(
      "region",
      "sub_region",
      "territory",
      "source",
      "org",
      "variable",
      "1995",
      "1996",
      "1997",
      "1998",
      "1999",
      "2000",
      "2001",
      "2002",
      "2003",
      "2004",
      "2005",
      "2006",
      "2007",
      "2008",
      "2009",
      "2010"
    )
  ) %>%
  slice(-(n()-14):-n())

## Repeatedly lag-fill until NA's are gone
countna <- function(df, v_vars) {
  df %>%
    filter_at(vars(!!!v_vars), any_vars(is.na(.))) %>%
    dim %>%
    .[[1]]
}

v_vars <- c("region", "sub_region", "territory", "source", "org")
while (countna(df_gun_raw, v_vars) > 0) {
  df_gun_raw <-
    df_gun_raw %>%
    mutate_at(vars(!!!v_vars), ~if_else(is.na(.), lag(.), .))
}

## Cast all the estimates
df_gun_raw <-
  df_gun_raw %>%
  mutate_at(
    vars(`1995`:`2010`),
    as.numeric
  )

## Reshape
df_gun_raw <-
  df_gun_raw %>%
  pivot_longer(
    `1995`:`2010`,
    names_to = "year",
    values_to = "value",
    names_ptypes = list(year = integer())
  ) %>%
  mutate(
    variable = case_when(
      str_detect(variable, "percentage") ~ "percent",
      str_detect(variable, "number") ~ "count",
      str_detect(variable, "rate") ~ "gun_rate",
      TRUE ~ variable
    )
  ) %>%
  pivot_wider(
    names_from = variable,
    values_from = "value"
  )

## Final cleaning
df_undoc_gun <-
  df_gun_raw %>%
  rename_all(str_to_lower)

df_undoc_gun %>%
  glimpse
```

Now we join with the general homicide data to visualize.

```{r gun-vis}
territories <-
  c(
    "United States of America",
    "Canada",
    "Germany",
    "Italy",
    "Spain"
  ) %>%
  paste(., collapse = "|")

full_join(
  df_undoc %>% select(territory, year, rate),
  df_undoc_gun %>% select(territory, year, gun_rate),
  by = c("territory", "year")
) %>%
  filter(
    str_detect(territory, territories),
    2003 <= year,
    year <= 2012
  ) %>%
  mutate(
    territory = if_else(str_detect(territory, "America"), "USA", territory)
  ) %>%

  ggplot(aes(year, fill = fct_reorder2(territory, year, rate))) +
  geom_bar(
    aes(y = rate, color = "General"),
    stat = "identity",
    position = position_dodge()
  ) +
  geom_bar(
    aes(y = gun_rate, color = "Firearm"),
    stat = "identity",
    position = position_dodge()
  ) +
  ## Tail
  scale_x_continuous(breaks = c(2000, 2003, 2006, 2009, 2012)) +
  scale_fill_discrete(name = "Country") +
  scale_color_manual(name = "Type", values = c("General" = NA, "Firearm" = "black")) +
  scale_linetype_manual(
    name = "Type",
    values = c("General" = 1, "Firearm" = 2)
  ) +
  theme_common() +
  labs(
    x = "Year",
    y = "Homicides per 100,000 Persons",
    caption = "Source: UNDOC Homicide Statistics"
  )
```

Observations:

- Firearm homicides account for a majority of US homicides; Italy has a similar trend.
- US *firearm* homicides occur at a greater rate than *general* homicides in similar nations.
- Data availability on firearm homicides is less than general homicide data.
