---
title: "Australia Fires"
author: "Zach"
date: 2019-01-11
output:
  github_document
---

Tuesday's exercise is investigating the Australia fires. The code below pulls the most recent fire reports, so the data are up-to-date as of knitting (2019-01-11).

```{r setup}
library(tidyverse)
library(sf)
library(lubridate)

# Get the Data

df_rainfall <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-07/rainfall.csv')
df_temperature <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-07/temperature.csv')

# IF YOU USE THIS DATA PLEASE BE CAUTIOUS WITH INTERPRETATION
df_nasa_fire <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-07/MODIS_C6_Australia_and_New_Zealand_7d.csv')

# For JSON File of fires
url <- "http://www.rfs.nsw.gov.au/feeds/majorIncidents.json"

df_aus_fires <- sf::st_read(url)
```

Download and unpack Australia boundary data.

```{r download}
url_aus_geo <- "http://data.daff.gov.au/data/warehouse/nsaasr9nnd_022/nsaasr9nnd_02211a04es_geo___.zip"
curl::curl_download(url_aus_geo, destfile = "../data/aus.shp.zip")
unzip("../data/aus.shp.zip", exdir = "../data")
```

Load boundary data.

```{r boundaries}
df_boundaries <- sf::read_sf("../data", "aust_cd66states")
```

Process the descriptions for `control` information.

```{r proc_control}
df_aus_control <- 
  df_aus_fires %>% 
  mutate(
    control = str_extract(description, "STATUS: [\\w|\\s]+") %>% 
      str_remove(., "STATUS: ") %>% 
      str_trim(.) %>% 
      ordered(., c("Out of control", "Being controlled", "Under control")),
    pubDate = dmy_hms(pubDate)
  )

df_aus_control %>% glimpse
```

```{r}
df_aus_control %>% pull(control) %>% unique
```

## Visualize
<!-- ----------------------------------------------------------------- -->

### GIS
<!-- --------------------------- -->

Geographic distribution of fires. This will give us valuable context to understand the other data.

```{r gis_vis}
fires_crs <- st_crs(df_aus_control)

# Lat-lon data for Australian cities
df_cities_latlon <- tribble(
       ~lat,     ~lon, ~city,
  +115.8605, -31.9505, "Perth",
  +138.6007, -34.9285, "Adelaide",
  +153.0251, -27.4698, "Brisbane",
  +151.2093, -33.8688, "Sydney",
  +149.1300, -35.2809, "Canberra",
  +144.9631, -37.8136, "Melbourne"
)
sf_cities_latlon <- 
  st_as_sf(
    df_cities_latlon,
    coords = c("lat", "lon"),
    crs = 4326,
    agr = "constant"
  )

ggplot() +
  geom_sf(data = df_boundaries %>% st_set_crs(fires_crs)) +
  geom_sf(
    data = df_aus_control,
    mapping = aes(fill = control, color = control)
  ) +
  geom_sf(
    data = sf_cities_latlon,
    shape = 21,
    size = 3,
    color = "red",
    fill = "black"
  ) +
  geom_text(
    data = df_cities_latlon,
    aes(lat, lon, label = city),
    hjust = 0,
    nudge_x = +0.5
  ) +
  ## Tail
  scale_fill_discrete(name = "Fire control status") +
  coord_sf(
    xlim = c(+142, +155),
    ylim = c(-38, -27)
  ) +
  guides(color = FALSE) +
  theme(legend.position = "right") +
  theme_minimal() +
  labs(x = "", y = "")
```

The fires are concentrated around Canberra and Sydney. Of course, smoke is probably affecting nearby cities as well.
The presence of "Out of control" fires near Canberra is cause for major alarm.

### Temperature
<!-- --------------------------- -->

What is the temperature history for Australia?

```{r}
df_temperature %>% glimpse
```


```{r vis-temperature}
df_temperature %>% 
  mutate(
    date = ymd(date),
    site_name = str_to_title(site_name)
  ) %>% 
  filter(date > ymd("2019-01-01")) %>%

  ggplot(aes(date, temperature, color = fct_reorder2(site_name, date, temperature))) +
  geom_point(
    data = . %>% mutate(flag = str_detect(site_name, "Canberra")),
    aes(alpha = flag),
    size = 0.5
  ) +
  geom_smooth(se = FALSE, size = 1) +
  
  scale_color_discrete(name = "Site") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 1 / 4)) +
  guides(alpha = FALSE)
```

The temperature near Canberra doesn't seem especially high; if anything, it's been colder than other parts of Australia.

### Rainfall
<!-- --------------------------- -->

```{r}
df_rainfall %>% glimpse
```

```{r vis-rainfall}
df_rainfall %>% 
  mutate(
    date = ymd(str_c(year, month, day))
  ) %>% 
  filter(date >= ymd("2019-01-01")) %>% 
  
  ggplot(aes(date, rainfall, color = fct_reorder2(city_name, date, rainfall))) +
  geom_point(
    data = . %>% mutate(flag = city_name == "Canberra"),
    aes(alpha = flag)
  ) +
  geom_smooth(se = FALSE) +
  
  scale_y_log10() +
  scale_color_discrete(name = "City") + 
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 1 / 4)) +
  guides(alpha = FALSE)
  
```

It seems July was dryer in Canberra than other parts of Australia. Perhaps this helped cause the fires?

The last few rainfall measurements in Canberra were also fairly low.
