---
title: "Marbles"
author: "Zach del Rosario"
date: 2020-06-02
output:
  github_document
---

Today's TidyTuesday is on ["Jelle's Marble Runs"](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-06-02).

```{r setup}
library(tidyverse)
library(rzdr)

df_marbles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-06-02/marbles.csv')
```

# First Look
<!-- -------------------------------------------------- -->

```{r }
df_marbles %>% glimpse()
```

```{r }
df_marbles %>% summary()
```

It's unclear to me how `avg_time_lap` is computed; what is the average over?

```{r }
df_marbles %>%
  mutate(
    time_comp = avg_time_lap * number_laps,
    time_diff = time_s - time_comp
  ) %>%
  summary
```

`time_comp` is approximately equal to `time_s`; it appears that `avg_time_lap` is the average over all laps.

# 1D EDA
<!-- -------------------------------------------------- -->

```{r vis-marble-times}
df_marbles %>%

  ggplot(
    aes(fct_reorder(marble_name, avg_time_lap, na.rm = TRUE), avg_time_lap, fill = team_name)
  ) +
  geom_boxplot() +

  scale_fill_discrete(name = "Team") +
  coord_flip() +
  theme_common() +
  labs(
    x = "Marble",
    y = "Average Lap Time (s)"
  )
```

**Observations**:

- There's certainly variation in the lap times, but it could easily be due to noise alone

```{r vis-track-times}
df_marbles %>%

  ggplot(aes(fct_reorder(site, avg_time_lap, na.rm = TRUE), avg_time_lap)) +
  geom_boxplot() +

  coord_flip() +
  theme_common() +
  labs(
    x = "Track",
    y = "Average Lap Time (s)"
  )
```

**Observations**:

- There is definitive variation in lap time with track

# Modeling
<!-- -------------------------------------------------- -->

### ANOVA
<!-- ------------------------- -->

Is there a significant difference in speeed among the marbles?

```{r anova-marbles-speed}
df_marbles %>%
  mutate(speed = avg_time_lap / track_length_m) %>%
  lm(speed ~ marble_name, data = .) %>%
  anova() %>%
  broom::tidy()
```

There's no strong effect of individual marble. For comparison, let's do the same
with sites.

```{r anova-track-time}
df_marbles %>%
  lm(avg_time_lap ~ site, data = .) %>%
  anova() %>%
  broom::tidy()
```

There's a huge effect of track site on lap time.

### Linear Models
<!-- ------------------------- -->

Can we predict `avg_time_lap` well based on `site`?

```{r lm-track-site}
df_marbles %>%
  lm(
    formula = avg_time_lap ~ site,
    data = .
  ) %>%
  summary()
```

The `site` variable carries a lot of information; how much can we recapture with
a coarser description of the track using just the features`number_laps` and
`track_length_m`.

```{r lm-track-feat}
df_marbles %>%
  lm(
    formula = avg_time_lap ~ number_laps + track_length_m,
    data = .
  ) %>%
  summary()
```

We get nearly the same COD using these features. This is nice because we use
fewer DOF for the fit, which will help with the model below:

Now let's fit on track features along with marble names; this will model the
variation in marble lap times, controlling for the track.

```{r lm-track-both}
df_marbles %>%
  lm(
    formula = avg_time_lap ~ number_laps + track_length_m + marble_name,
    data = .
  ) %>%
  anova() %>%
  broom::tidy()
```

Once again, we see that marble identity does not have a significant effect on lap time.
