---
title: "US PhD's Awarded: EDA"
author: "Zachary del Rosario"
date: 2019-11-08
output:
  github_document
---

Exploration of "PhD's awarded data" from the NSF. Inspired by Tidytuesday.

Sources:

- Original: [NSF](https://ncses.nsf.gov/pubs/nsf19301/data)
- Via: [Tidytuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-19)

## Setup
<!-- -------------------------------------------------- -->

```{r setup}
library(tidyverse)

phd_field <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")
```

Some overview checks.

```{r overview}
phd_field %>% glimpse

phd_field %>% summary
```

What are the major fields coded as?

```{r major_fields}
phd_field %>%
  select(broad_field, major_field) %>%
  unique %>%
  print(n = 25)
```

These are not the most informative broad fields. I will do some re-coding, then
down-select to some broad areas of interest.

```{r fields}
df_recoded <-
  phd_field %>%
  mutate(
    broad_field = case_when(
      str_detect(major_field, "Physics") ~ "Physical sciences",
      str_detect(major_field, "Chemistry") ~ "Physical sciences",
      str_detect(major_field, "Geosciences") ~ "Physical sciences",

      str_detect(major_field, "Computer") ~ "Computer science",
      str_detect(major_field, "Mathematics") ~ "Math and Stats",
      TRUE ~ broad_field
    )
  )

broad_fields_selected <-
  c(
    "Physical sciences",
    "Computer science",
    "Math and Stats",
    "Engineering"
  )
```


## Visualizations
<!-- -------------------------------------------------- -->

How many sub-disciplines (levels of `field`) are in each major field?

```{r count-fields}
df_recoded %>%
  select(broad_field, major_field, field) %>%
  unique %>%
  count(broad_field, major_field) %>%
  group_by(broad_field) %>%
  mutate(n_max = max(n)) %>%
  ungroup() %>%
  mutate(broad_field = fct_reorder(broad_field, -n_max)) %>%

  ggplot(aes(fct_reorder(major_field, n), n)) +
  geom_col() +
  facet_grid(
    broad_field~.,
    scales = "free_y",
    labeller = labeller(broad_field = label_wrap_gen(width = 10))
  ) +
  coord_flip() +
  theme(strip.text.y = element_text(angle = 0)) +
  labs(
    x = "Major Fields",
    y = "Sub-disciplines"
  )
```

Observations:

- The `Life Sciences`; particularly biosciences, have many, many sub-disciplines.
- Disciplines like `Economics`, `Anthropology`, and `CS` have very few sub-disciplines.

## Timeseries
<!-- -------------------------------------------------- -->

By broad field, how many PhD's are minted?

```{r summarize-fields}
df_recoded %>%
  group_by(broad_field, year) %>%
  summarize(n_phds = sum(n_phds, na.rm = TRUE)) %>%
  ungroup() %>%

  ggplot(aes(year, n_phds, color = fct_reorder2(broad_field, year, n_phds))) +
  geom_line() +
  scale_x_continuous(breaks = seq(from = 2008, to = 2016, by = 2)) +
  scale_y_log10() +
  scale_color_discrete(name = "Broad Field") +
  labs(
    x = "Year",
    y = "PhDs"
  )
```

Observations:

- I had no idea there were so many PhD's earned in the Social sciences!
- There was a fairly sharp decline in Education post-Great Recession (2008),
  which has not recovered.
- CS, Math and Stats, and (especially) Engineering saw an increase from 2010 to
  2014, but these seem to have leveled out.

Let's disaggregate by sub-discipline and inspect the timeseries again.

```{r selected-timeseries}
df_recoded %>%
  filter(broad_field %in% broad_fields_selected) %>%
  group_by(field) %>%
  mutate(delta = lead(n_phds) - n_phds) %>%
  ungroup() %>%
  group_by(broad_field, year) %>%
  mutate(n_phds_total = sum(n_phds, na.rm = TRUE)) %>%
  ungroup() %>%

  ggplot(aes(year, n_phds, group = field)) +
  geom_line(aes(color = delta > 0), show.legend = FALSE) +
  geom_line(aes(y = n_phds_total), color = "grey80") +
  facet_grid(~broad_field) +
  scale_y_log10() +
  scale_x_continuous(breaks = c(2008, 2012, 2016)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Year",
    y = "PhDs"
  )
```

Observations:

- The number of PhD's awarded in the Computer science(s), Math and Stats, and
  Engineering are roughly equal. There are many more PhD's awared in the
  Physical sciences.
- A certain Engineering sub-discipline seems to have declined and then
  disappeared in 2014.
- There was a fairly sharp decline across various Math and Stats sub-disciplines
  between 2014-16.

Build a helper function to highlight the top and bottom sub-disciplines.

```{r timeseries-helper}
plot_timeseries <- function(df) {
  fields_minmax <-
    df %>%
    group_by(field) %>%
    summarize(
      n_min = min(n_phds, na.rm = TRUE),
      n_max = max(n_phds, na.rm = TRUE)
    ) %>%
    filter((dense_rank(n_min) <= 5) | (dense_rank(-n_max) <= 5)) %>%
    pull(field)

  df %>%
  ggplot(aes(year, n_phds)) +
  geom_line(aes(group = field), color = "grey80") +
  geom_line(
    data = . %>% filter(field %in% fields_minmax),
    aes(color = fct_reorder2(field, year, n_phds))
  ) +
  scale_y_log10() +
  scale_x_continuous(breaks = c(2008, 2012, 2016)) +
  scale_color_discrete(name = "Field") +
  guides(color = guide_legend(ncol = 1)) +
  theme(legend.position = "right") +
  labs(
    x = "Year",
    y = "PhDs"
  )
}
```

```{r cs-timeseries}
df_recoded %>%
  filter(broad_field == "Computer science") %>%
  plot_timeseries
```

Pure CS degrees seem most common; I'm not sure what the distinction is with
other degrees.

```{r mathstat-timeseries}
df_recoded %>%
  filter(broad_field == "Math and Stats") %>%
  plot_timeseries
```

Observations:

- Statistics seems to be the most popular sub-discipline, especially if we
  consider the `Mathematics and statistics` flavors as the same. Applied math is
  quite high as well.
- The "crash" we saw earlier seems to be confined to pure disciplines.
- There are very few *Mathematics* OR PhD's trained every year!

```{r engr-timeseries}
df_recoded %>%
  filter(broad_field == "Engineering") %>%
  plot_timeseries
```

Observations:

- There are far more OR PhD's classified in Engineering (rather than
  Mathematics).

```{r phys-timeseries}
df_recoded %>%
  filter(broad_field == "Physical sciences") %>%
  plot_timeseries
```
