---
title: "IMDB Ratings"
author: "Zach"
date: 2019-12-31
output:
  github_document
---

Since there's no tidytuesday this week, I went back to the archives and decided
to try the [IMDB
dataset](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-01-08)
posted earlier this year.

Some setup:

```{r setup}
library(tidyverse)
library(lubridate)

df_imdb <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-08/IMDb_Economist_tv_ratings.csv")
```

## Simple EDA
<!-- -------------------------------------------------- -->

```{r }
df_imdb %>% glimpse
```

```{r }
df_imdb %>%
  arrange(desc(av_rating)) %>%
  select(av_rating, title, everything()) %>%
  head
```

Are shows rated on a per-season basis?

```{r }
df_imdb %>%
  filter(title == "Breaking Bad")
```

Yes, it appears each show is listed with all of its seasons. I'm not sure how to
read these dates; "Breaking Bad" Season 5
[half-premiered](https://en.wikipedia.org/wiki/Breaking_Bad#Season_5_(2012%E2%80%9313))
on 2013-08-11, not 2013-02-20. Given the "Polizeiruf 110" data below, these dates
must be connected to ratings, and not the airing dates themselves.

Which shows have the most seasons?

```{r }
df_imdb %>%
  group_by(titleId) %>%
  mutate(seasons = max(seasonNumber)) %>%
  ungroup() %>%
  filter(seasons == seasonNumber) %>%
  arrange(desc(seasons)) %>%
  select(title, seasons, av_rating, date) %>%
  head(10)
```

Note that since this is a *ratings* databse, we won't catch shows that haven't
been rated. That being said [Polizeiruf
110](https://en.wikipedia.org/wiki/Polizeiruf_110) has been running since 1971,
first in the German Democratic Republic (East Germany).

Why isn't "The Simpsons" in this database? That's weird.

What's the distribution of seasons among shows?

```{r season-hist}
df_imdb %>%
  group_by(titleId) %>%
  summarize(seasons = max(seasonNumber)) %>%
  count(seasons)
```

A very large fraction of shows have just one season; fewer are renewed for a
second season, and a modest number continue for three or more.

The genres are given as lists; let's split those to make the genre data parsable.

```{r }
df_genres <-
  df_imdb %>%
  mutate(genre = str_split(genres, ",")) %>%
  unnest(genre)

df_genres %>%
  count(genre) %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

I have a hunch this distribution is more a data artifact than anything else; I
feel like I can name more than two Reality-TV shows!

## Visualize
<!-- -------------------------------------------------- -->

Let's take a rough look at all the data over time:

```{r }
df_imdb %>%

  ggplot(aes(date, av_rating)) +
  geom_point(aes(color = titleId), show.legend = FALSE) +
  geom_line(aes(color = titleId), show.legend = FALSE) +
  geom_smooth()
```

I connect shows across seasons, and color by show ID.

Observations:

- There are a lot more shows later in time than earlier. This is probably in large part a data artifact (IMDB was [launched](https://en.wikipedia.org/wiki/IMDb) in 1990), though there very well may be more TV shows today given the rise of streaming platforms.
- I don't see an especially large trend in average ratings, aside from a bit of a dip in the late 90's.

Do shows tend to trend up, down, or stay the same across seasons?

```{r season-trend}
df_imdb %>%

  ggplot(aes(seasonNumber, av_rating)) +
  geom_line(aes(color = titleId), show.legend = FALSE) +
  geom_smooth() +
  ## Tail
  coord_cartesian(xlim = c(0, 20))
```

Observations:

- Edge-effects on regression mean the downward trend near 20 is unreliable.
- There seems to be a "bump" in ratings around season 4-6, but...
- The large number of shows unrenewed for a 2nd season probably skew the trend; let's try again with longer-running shows.

```{r season-trend-filt}
df_imdb %>%
  group_by(titleId) %>%
  filter(max(seasonNumber) > 3) %>%
  ungroup() %>%

  ggplot(aes(seasonNumber, av_rating)) +
  geom_line(aes(color = titleId), show.legend = FALSE) +
  geom_smooth() +
  ## Tail
  coord_cartesian(xlim = c(0, 20))
```

Observations:

- The bump persists even in the filtered data, but I'm worried about edge effects pulling down the early-season part of the trend.

Let's try a boxplot series, which will not make structural assumptions on the data.

```{r season-trend-boxplots}
df_imdb %>%
  group_by(titleId) %>%
  filter(max(seasonNumber) > 3) %>%
  ungroup() %>%
  mutate(season = as.factor(seasonNumber)) %>%

  ggplot(aes(season, av_rating)) +
  geom_boxplot()
```

The bump does seem to be real!

I've felt for awhile that TV shows "take awhile" to get good; it seems that
trend is (weakly) reflected in the data!

Is there a difference in first-season ratings between shows that are and are not renewed?

```{r }
df_imdb %>%
  group_by(titleId) %>%
  mutate(renewed = (max(seasonNumber) > 1)) %>%
  ungroup() %>%
  filter(seasonNumber == 1) %>%

  ggplot(aes(renewed, av_rating)) +
  geom_boxplot()
```

There is a difference in ratings, but it's not huge and there's a fair bit of
overlap between the two categories.

Are there rating differences across genres?

```{r }
df_genres %>%
  group_by(genre) %>%
  ## filter(n() > 100) %>%
  mutate(mid_rating = median(av_rating, na.rm = TRUE)) %>%
  ungroup() %>%

  ggplot(aes(fct_reorder(genre, mid_rating), av_rating)) +
  geom_boxplot() +
  coord_flip()
```

The low sample size for Documentary and Reality shows means we can't draw strong
conclusions for those genres. Otherwise there's not an especially strong trend
across genres, though Western, Animation, and Sci-Fi tend to be lowest-rated and
War, Sport, and History highest-rated.

Are there differences in number of seasons across genres?

```{r }
df_genres %>%
  group_by(titleId) %>%
  mutate(seasons = max(seasonNumber)) %>%
  ungroup() %>%
  group_by(genre) %>%
  mutate(mid_season = median(seasons, na.rm = TRUE)) %>%
  ungroup() %>%

  ggplot(aes(fct_reorder(genre, mid_season), seasons)) +
  geom_boxplot() +
  coord_flip()
```

Observations:

- We've already seen that there are very few Musical, Documentary, and Reality shows in the dataset, so those statistics are not generalizable.
- Certain genres seem to have long-running outlier shows, such as Drama, Crime, and Adventure.
