---
title: ""
author: ""
date:
output:
  github_document
---

```{r setup}
library(tidyverse)

df_food <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
```

```{r theme-setup}
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title = element_text(size = 18),
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12)
  )
}

```

```{r summary}
df_food %>% summary
```

```{r countries}
df_food %>%
  pull(country) %>%
  unique
```

```{r categories}
df_food %>%
  pull(food_category) %>%
  unique
```

```{r food-consumption}
df_food %>%
  group_by(food_category) %>%
  mutate(spread = sd(consumption)) %>%
  ungroup() %>%
  mutate(food_category = fct_reorder(food_category, spread)) %>%

  ggplot(aes(consumption)) +
  geom_histogram() +
  facet_wrap(~food_category)
```

Observations:

- Consumption of milk and cheese varies widely across countries
- Consumption of rice and wheat varies a fair bit

```{r emission-v-consumption}
df_food %>%
  ggplot(aes(consumption, co2_emmission)) +
  geom_point(aes(color = food_category))
```

Observations:

- CO2 emission is linear in consumption; this tells us either that 1. no country
  is significantly more efficient in their CO2 generation (the values are real),
  or 2. The emission data are calculated from consumption values (the values are
  estimated / synthetic)

```{r food-emission}
df_food %>%
  mutate(emission = co2_emmission / consumption) %>%
  group_by(food_category) %>%
  summarize(emission = mean(emission, na.rm = TRUE)) %>%

  ggplot(aes(fct_reorder(food_category, emission), emission)) +
  geom_point() +
  coord_flip()
```

Observations:

- Lamb, goat, and beef are by far the largest producers of CO2 emissions

```{r country-emission, fig.height=14}
df_food %>%
  group_by(country) %>%
  mutate(co2_total = sum(co2_emmission, na.rm = TRUE)) %>%
  ungroup() %>%

  ggplot(aes(fct_reorder(country, co2_total), co2_emmission)) +
  geom_bar(
    aes(fill = food_category),
    color = "black",
    stat = "identity"
  ) +

  scale_fill_discrete(name = "Category") +
  scale_y_continuous(breaks = c(0, 1000, 2000)) +
  coord_flip() +
  theme_common() +
  theme(
    axis.text.y = element_text(size = 8, hjust = 1),
    aspect.ratio = 5 / 1
  ) +
  labs(
    x = "Country",
    y = "Emission (Kg CO2 / person / yr)"
  )
```

Observations:

- The top producers of CO2 are Argentina, Australia, Albania, New Zealand, and
  Iceland.
  + There's variety in how these countries produce CO2; Argentina produces
    emissions through beef production, while Iceland produces primarily through
    lamb and goat
