---
title: "School Shootings"
author: "Zach"
date: 2020-01-20
output:
  github_document
---

I found these data by reading Andrew Gelman's
[blog](https://statmodeling.stat.columbia.edu/2020/01/18/graphs-of-school-shootings-in-the-u-s/),
but the source is [CNN
Labs](https://github.com/cnnlabs/cnn-school-shooting-data/).

```{r setup}
library(tidyverse)
library(lubridate)
library(sf)

df_data <- readr::read_csv('https://raw.githubusercontent.com/cnnlabs/cnn-school-shooting-data/master/dataset.csv') %>%
  mutate(date = dmy(date))
```

## GIS Download
<!-- -------------------------------------------------- -->

```{r}
url_boundaries <- "https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_state_20m.zip"
file_boundaries = "../data/usa_states.zip"

curl::curl_download(url_boundaries, destfile = file_boundaries)
unzip(file_boundaries, exdir = "../data")

sf_boundaries <- sf::read_sf("../data", "cb_2018_us_state_20m")
```

## EDA
<!-- -------------------------------------------------- -->

What are the types of shootings?

```{r types}
df_data %>%
  count(type) %>%
  arrange(desc(n)) %>% 
  knitr::kable()
```

High school at the top conforms to my expectations, but Elementary school is
alarming. These are almost surely external perpretrators, while High school
probably involves more student actors.

Having all these types is accurate, but distracting. Let's aggregate some of the types. I choose to aggregate to the highest grade (e.g. K-12 -> High School). 

```{r aggregate}
df_agg <- 
  df_data %>% 
  mutate(
    class = case_when(
      type == "High School" ~ "High School",
      type == "Elementary School" ~ "Elementary School",
      type == "Middle School" ~ "Middle School",
      
      type == "Grades 7-12" ~ "High School",
      type == "K-12" ~ "High School",
      type == "Grades 7-8" ~ "Middle School",
      type == "K-8" ~ "Middle School",
      type == "PreK-12" ~ "High School",
      type == "PreK-8" ~ "Middle School"
    ),
    class = fct_relevel(class, "Elementary School", "Middle School", "High School")
  )
df_agg %>% glimpse
```


```{r victims}
df_agg %>%
  count(victims) %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

It's cold comfort, but at least the majority of shootings seem to have just one
victim.

## Timeseries
<!-- -------------------------------------------------- -->

```{r timeseries}
df_agg %>%

  ggplot(aes(date, victims)) +
  geom_point(aes(color = class)) +
  geom_text(
    data = . %>% filter(victims > 10),
    aes(label = school),
    hjust = 1,
    nudge_y = 0.05
  ) +

  scale_y_log10() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Observations:

- Many of the "mass" school shootings have occurred in recent times
- It's hard to see from this vis, but the number of single-victim shootings seems to have increased in recent times as well.

CNN produced a bargraph that illustrates the count of shootings fairly well; let's reproduce that:

```{r }
df_agg %>% 
  mutate(year = as.factor(year)) %>% 
  
  ggplot(aes(year, fill = class)) +
  geom_histogram(stat = "count")
```

This does illustrate the upward trend in shootings. With the previous figure, we can see that in recent times there have been more frequent and more deadly shootings.

## Times of day
<!-- -------------------------------------------------- -->

One of the figures Gelman discusses is meant to illustrate that more school
shootings occur on Friday afternoons. The original figure is hard to interpret.
Can we do better?

```{r times-v-weekday}
df_data %>%
  mutate(day = wday(date, label = TRUE)) %>%

  ggplot(aes(day, time)) +
  geom_point(
    aes(fill = year),
    pch = 21, 
    color = "grey50", 
    position = position_jitter(width = 0.2, height = 0)
  ) +

  viridis::scale_fill_viridis() +
  scale_y_reverse(labels = function(time) {
    str_c(round(time / 3600), ":00")
  }) +
  theme_minimal()
```

Observations:

- There does not seem to be a pattern in year and time; the shootings seem to take place in roughly the same times & days over the years
- As CNN notes, there seem to be more shootings on Friday afternoons. But there are also a fair number on Friday mornings.

Friday in general seems to be more common:

```{r weekdays}
df_data %>%
  mutate(day = wday(date, label = TRUE)) %>%

  ggplot(aes(day)) +
  geom_histogram(stat = "count") +

  theme_minimal()
```

Of schooldays, Monday is least common and Friday most common.

## Geospatial
<!-- -------------------------------------------------- -->

Let's process the geospatial data into simple feature format for plotting.

```{r gis-proc}
sf_data <- 
  st_as_sf(
    df_agg %>% filter(!is.na(LONCOD), !is.na(LATCOD)), 
    coords = c("LONCOD", "LATCOD"),
    crs = 4326, 
    agr = "constant"
  )

sf_data %>% glimpse
```

Now let's visualize.

```{r gis-vis}
sf_data %>% 
  ggplot() +
  geom_sf(
    data = sf_boundaries
  ) +
  geom_sf(
    aes(fill = year),
    pch = 21,
    size = 1,
    color = "black"
  ) +
  
  viridis::scale_fill_viridis() +
  coord_sf(
    xlim = c(-130, -60),
    ylim = c(+25, +50)
  ) +
  theme_minimal()
```

Observations:

- I don't see any clear pattern, other than the intuitive observation that shootings seem to take place in locations of fairly high population density (fewer in the upper-west)
